---
title: "PS5 Rmarkdown"
author: "Cobi Stancik"
date: "2023-02-20"
output: html_document
---
**Load data...**
```{r, message=FALSE}
library(tidyverse)
gapminder <- read_delim("gapminder.csv.bz2")
```

**How many rows/columns do we have?**
```{r}
gapminder %>%
  dim()
```

**Printing a small sample of data..**
```{r}
gapminder %>%
  sample_n(10)
```

**How many countries are there in the dataset?**
```{r}
gapminder %>%
  summarize(
    iso2_countries = n_distinct(iso2),
    iso3_countries = n_distinct(iso3),
    name_countries = n_distinct(name)
  )
```

**If you did this correctly, you saw that there are more names than iso-2 codes, and there are even more is03-codes.** 
**What is going on? Can you find it out?**

Took awhile but its because some countries have separated splitting into two countries, which I should have realized earlier since I am actually like 60% Czech, so with Czechoslovakia it used to be a single country with the ISO-2 code "CS", but it split into the Czech Republic ("CZ") and Slovakia ("SK") in 1993.

**Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?**
```{r}
gapminder %>%
  group_by(iso2, name) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  filter(n > 1) %>% 
  count()
```

**Now repeat the same for name and is03-code. Are there country names that have more than one is03-code? What are these countries?**
```{r}
gapminder %>%
  group_by(iso3, name) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  filter(n > 1) %>% 
  count()
```

**What is the minimum and maximum year in these data?**
```{r}
gapminder %>%
  summarize(
    min_year = min(time, na.rm = TRUE),
    max_year = max(time, na.rm = TRUE)
  )
```


**How many missing c02 emissions are there for each year? Analyze both missing C02 and c02_PC. Which years have most missing data?**
```{r}
gapminder %>%
  group_by(time) %>%
  summarize(
    missing_co2 = sum(is.na(co2)),
    missing_co2_pc = sum(is.na(co2_PC))
  ) %>%
  arrange(desc(missing_co2))
```

**Make a plot of total C02 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.**

We can see that while the US has consistently been the largest emitter of CO2, China has overtaken it in recent years. India has also experienced a significant increase in CO2 emissions in the last few decades.
```{r}
gapminder %>%
  filter(iso3 %in% c("USA", "IND", "CHN", "JPN","DEU", "BRA")) %>% 
  ggplot +
  geom_line(aes(x = time, y = co2, color = iso3)) +
  geom_point(aes(x = time, y = co2, color = iso3))
```

**Now let's analyze the C02 emissions per capita. Make a similar plot of the same countries. What does this figure suggest?**

We can see that while the US and Japan have had relatively stable CO2 emissions per capita, China and India have experienced a significant increase in CO2 emissions per capita in the last few decades.
```{r}
gapminder %>%
  filter(iso3 %in% c("USA", "IND", "CHN", "JPN", "DEU", "BRA")) %>% 
  ggplot +
  geom_line(aes(x = time, y = co2_PC, color = iso3)) +
  geom_point(aes(x = time, y = co2_PC, color = iso3))
```

**Compute average C02 emissions per capita across the continents (assume region is the same as continent). Comment what do you see.**

We can see that while all continents have experienced an increase in CO2 emissions per capita, North America has consistently had the highest emission
```{r}
gapminder %>%
  group_by(region, time) %>%
  summarize(avg_co2_pc = mean(co2_PC, na.rm = TRUE)) %>%
  filter(time %in% c(1960, 2016))
```


Make a barplot where you show the previous results—average C02 emissions per capita across continents in 1960 and 2016.
```{r}
avg_co2_pc <- gapminder %>%
  group_by(region, time) %>%
  summarize(avg_co2_pc = mean(co2_PC, na.rm = TRUE)) %>%
  filter(time %in% c(1960, 2016))

ggplot(avg_co2_pc, aes(x = region, y = avg_co2_pc, fill = factor(time))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average CO2 Emissions per Capita Across Continents",
       x = "Continent",
       y = "Average CO2 Emissions per Capita",
       fill = "Time")
```

**Which countries are the three largest, and three smallest C02 emitters (in terms of C02 percapita) in 2019 for each continent? (Assume region is continent)**

**Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960, Comment what do you see there.**

We can see that there is a positive relationship between GDP per capita and life expectancy, as countries with higher GDP per capita tend to have higher life expectancies.

```{r}
gdp_PC <- gapminder %>% 
  filter(time == 1960) %>% 
  select(region, GDP_PC, lifeExpectancy, totalPopulation)

ggplot(gdp_PC, aes(x = GDP_PC, y = lifeExpectancy, color = factor(region))) +
  geom_point(aes(size = totalPopulation)) +
  scale_size(range = c(1, 15)) +
  labs(title = "GDP per capita vs. Life Expectancy by Country (1960)",
       x = "GDP per capita",
       y = "Life Expectancy",
       color = "Continent")
```

**Make a similar plot, but this time use 2019 data only.**

```{r}
gdp_PC <- gapminder %>% 
  filter(time == 2019) %>% 
  select(region, GDP_PC, lifeExpectancy, totalPopulation)

ggplot(gdp_PC, aes(x = GDP_PC, y = lifeExpectancy, color = factor(region))) +
  geom_point(aes(size = totalPopulation)) +
  scale_size(range = c(1, 15)) +
  labs(title = "GDP per capita vs. Life Expectancy by Country (2019)",
       x = "GDP per capita",
       y = "Life Expectancy",
       color = "Continent")
```

**Compare these two plots and comment what do you see. How has world developed through the last 60 years?**

Overall, the plots suggest that the world has experienced significant development and progress in terms of both economic and human development over the last 60 years.

**Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?**

```{r}
# Average life expectancy by continent in 1960
gapminder %>% 
  filter(time == 1960) %>% 
  group_by(region) %>% 
  summarize(avg_lifeExp = mean(lifeExpectancy, na.rm = TRUE))

# Average life expectancy by continent in 2019
gapminder %>% 
  filter(time == 2019) %>% 
  group_by(region) %>% 
  summarize(avg_lifeExp = mean(lifeExpectancy, na.rm = TRUE))
```


**Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see**

Africa had the highest average LE growth over the past 60 years, followed by Asia, Oceania, Americas, and Europe.

```{r}
# Average life expectancy growth by continent from 1960-2019
gapminder %>% 
  group_by(region) %>% 
  summarize(avg_growth = (mean(lifeExpectancy[time == 2019], na.rm = TRUE) - 
                          mean(lifeExpectancy[time == 1960], na.rm = TRUE)) / 
                          mean(lifeExpectancy[time == 1960], na.rm = TRUE)) %>%
  arrange(desc(avg_growth))
```

**Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it!**

```{r}
gapminder %>% 
  filter(time %in% c(1960, 2019)) %>%
  ggplot(aes(x = GDP_PC, fill = factor(time))) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  ggtitle("Histogram of GDP per capita (1960 vs. 2019)") +
  xlab("GDP per capita") + ylab("Count") +
  scale_fill_discrete(name = "Year") 
```

**What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top)**
```{r}
ranks_per_year <- gapminder %>%
  select(name, lifeExpectancy, time) %>%
  filter(!is.na(lifeExpectancy), !is.na(name), time %in% c("1960", "2019")) %>% 
  arrange(time, desc(lifeExpectancy)) %>% 
  group_by(time) %>% 
  mutate(rankings = rank(desc(lifeExpectancy)))

ranks_per_year %>% 
 filter(name == "United States")
```

Very tired probably ~10hours :)